On hyperparameter tuning through Lipschitz global optimization

### 2024-03-21 2024/ITMO paper/sn-article.tex
Add new Results section

%% Русский язык - закомментировать в финальной версии
\usepackage[russian]{babel}
%% Зачеркивание - закомментировать в финальной версии
\usepackage[normalem]{ulem}

@@

\abstract{The quality of machine learning methods is substantially affected by their hyperparameters, while the evaluation of the quality criterion is a time-consuming operation. Therefore, it is important to develop intelligent methods for selecting optimal values of hyperparameters that require a small number of search trials. In this paper, we propose a new approach to hyperparameter tuning based on the ideas of Lipschitz global optimization. In the framework of this approach, the solution of problems with several parameters is reduced to solving equivalent one-dimensional problems. The reduction is based on the use of space-filling curves (Peano curves). These approaches are implemented in the open-source framework of \textit{i}ntelligent \textit{opt}imization methods \textcolor[rgb]{1,0,0}{(iOpt)}.  To demonstrate the advantages of iOpt, we compare it with the well-known Optuna and HyperOpt frameworks when tuning hyperparameters of various machine learning methods on \textcolor[rgb]{1,0,0}{\sout{a} the}  representative \textcolor[rgb]{1,0,0}{ \sout{set of} } datasets. The results show that Lipschitz global optimization methods provide comparable (in terms of quality) results in a significantly shorter time compared to known hyperparameter tuning algorithms.}

@@

In this paper, we present a framework of \textit{i}ntelligent \textit{opt}imization methods  \textcolor[rgb]{1,0,0}{iOpt\footnote{\url{https://github.com/aimclub/iOpt/}}}, focused on solving problems of selecting optimal values of hyperparameters of various algorithms (including machine learning \textcolor[rgb]{1,0,0}{and artificial intelligence} methods).

\textcolor[rgb]{1,0,0}{Отличительной особенностью фреймворка iOpt является использование for hyperparameters tuning методов липшицевой оптимизации, которые являются детерминированными, обладают быстрой сходимостью, а также возможностью адаптации к различного рода специальным постановкам задач.
Например, рассматриваемые методы допускают обобщение для работы с зашумленными функциями \citep{Sergeyev2020}. В данной работе они были адаптированы для решения HPO problems, в которых часть параметров часто являются дискретными и/или категориальными.}

@@

%Новый текст на русском
\textcolor[rgb]{1,0,0}{
Условия сходимости Global Search Algorithm для задач с Categorical Parameters (GSA-CP) могут быть сформулированы в виде следующего утверждения.
\begin{proposition}
Пусть функции $ \varphi(y,u_s), y \in D, s = 1,...,S,$ form (\ref{problem_is}) удовлетворяют условию Липшица с соответствующими константами $L_s, s = 1,...,S$, и $L = \max\{L_s\}$.
И пусть точка $\overline{x}$ be a limit point of the sequence $\left\{x^k\right\} $, полученной при минимизации функции $f(x)$ из (\ref{f}) с помощью алгоритма GSA-CP. Then:
\begin{enumerate}
	\item 
	если одноверменно с $\overline{x}$ существует другая предельная точка $x'$ последовательности $\left\{x^k\right\}$, тогда $f(\overline{x}) = f(x')$.
	\item
	$z^k=f(x^k) \geq f(\overline{x})$ при любом $k \geq 0$.
	\item
	Если на некотором шаге поиска величина $\mu$ from (\ref{mu}) satisfies the condition
\begin{equation}\label{217}
  r \mu > 2^{3-1/N} L \sqrt{N+3},
\end{equation}
then $\overline{x}$ is a global minimizer of the function $f(x), x\in[0,S]$ and any global minimizer $x^\ast$ is also a limit point of the sequence $\left\{x^k\right\}$.
\end{enumerate}
\end{proposition}
\begin{proof}
Это утверждение доказывается аналогично теореме о сходимости index method of global optimization from \citet{Strongin2000}, принимая во внимание взаимосвясь (\ref{f}) функций  $f(x)$ и $\varphi(y,u_s)$, а также правила (\ref{index}), задающие индексы у точек испытаний.
\end{proof}
}

@@

\subsection{Results and discussion}

@@

Дополнительная серия экспериментов по сравнению iOpt, Optuna и HyperOpt была проведена с использованием фреймворка JAHS-Bench-201 [ссылка]. JAHS-Bench предоставляет удобный доступ к набору задач для определения оптимального сочетания гиперпараметров и архитектуры neural network. Данный инструмент для сравнения HPO-алгоритмов содержит обширный набор данных о качестве и производительности нейронных сетей, состоящий примерно из 161 million data points, каждая из которых состоит из 20 показателей.

@@

Особенностью JAHS-Bench является использование surrogate models fit on performance data, covering different optimization problems (single/multiple objectives), что предоставляет возможность варьировать архитектуру нейросети (выбирая один из имеющихся вариантов), а также ее гиперпараметры. При этом использование surrogate models позволяет оценить качество работы обученного классификатора при конкретном наборе гиперпараметров за доли секунды вместо нескольких часов, которые были бы затрачены на обучение реальной нейросети. В качестве основы для суррогатных моделей в JAHS-Bench используются три популярных датасета -- CIFAR-10, Colorectal-Histology, Fashion MNIST -- применяемых для задач классификации.

В проведенных экспериментах настройка гиперпараметров с помощью фреймворков iOpt, Optuna и HyperOpt выполнялась на фиксированной архитектуре нейросети, предложенной в JAHS-Bench качестве архитектуры по умолчанию. Для настройки были выбраны традиционно наиболее значимые (с точки зрения исследователя) гиперпараметры: категориальный параметр выбора функции активации (принимающий три возможных значения ReLU, Mish, Hardswish), а также непрерывные параметры Learning Rate and Weight Decay в диапазонах $[10^{-3},1]$ and $[10^{-5},10^{-2}]$ соответственно.

В качестве целевой функции использовалась метрика validation accuracy, вычисленная после 200 эпох обучения. Точность поиска в критерии остановки всех методов была установлена равной 0.01, а максимальное число испытаний – равным 200. Для алгоритмов из Optuna и HyperOpt достигнутые показатели усреднялись по 10 запускам. Результаты сравнения представлены в таблице \ref{tab:met-surr}. Как и ранее, the values corresponding to the best result on average are highlighted in bold and the values corresponding to the best result for the worst case are highlighted in red.

Результаты показывают превосходство iOpt на датасетах CIFAR-10 и Colorectal-Histology. При использовании датасета Fashion MNIST фреймворк iOpt превосходит HyperOpt, но уступает Optuna. Отметим, что при использовании суррогатных моделей время вычисления целевой функции составляет доли секунды, поэтому результаты, связанные со сравнением времени работы алгоритмов, для этого эксперимента не приводятся.

>>>>

%% Русский язык - закомментировать в финальной версии
%\usepackage[russian]{babel}
%% Зачеркивание - закомментировать в финальной версии
%\usepackage[normalem]{ulem}

@@

\abstract{The quality of machine learning methods is substantially affected by their hyperparameters, while the evaluation of the quality criterion is a time-consuming operation. Therefore, it is important to develop intelligent methods for selecting optimal values of hyperparameters that require a small number of search trials. In this paper, we propose a new approach to hyperparameter tuning based on the ideas of Lipschitz global optimization. In the framework of this approach, the solution of problems with several parameters is reduced to solving equivalent one-dimensional problems. The reduction is based on the use of space-filling curves (Peano curves). These approaches are implemented in the open-source framework of \textit{i}ntelligent \textit{opt}imization methods (iOpt).  To demonstrate the advantages of iOpt, we compare it with the well-known Optuna and HyperOpt frameworks when tuning hyperparameters of various machine learning methods on representative datasets. The results show that Lipschitz global optimization methods provide comparable (in terms of quality) results in a significantly shorter time compared to known hyperparameter tuning algorithms.}

@@

In this paper, we present a framework of \textit{i}ntelligent \textit{opt}imization methods  iOpt\footnote{\url{https://github.com/aimclub/iOpt/}}, focused on solving problems of selecting optimal values of hyperparameters of various algorithms (including machine learning and artificial intelligence methods).

An essential feature of the iOpt framework is the use of Lipschitz optimisation methods for hyperparameter tuning. These methods are deterministic, provide fast convergence, and can be adapted to various types of special problem formulations. For example, such methods allow generalisation for working with noisy functions  \citep{Sergeyev2020}. In this paper, these methods have been adapted to solve HPO problems where some of the parameters are often discrete and/or categorical.

@@

%Новый текст на русском
The convergence conditions of the Global Search Algorithm for problems with Categorical Parameters (GSA-CP) can be formulated as the following statement.
\begin{proposition}
Let the functions $ \varphi(y,u_s), y \in D, s = 1,...,S,$ from (\ref{problem_is}) satisfy the Lipschitz condition with respective constants $L_s, s = 1,...,S$, and $L = \max\{L_s\}$.
Also let the point $\overline{x}$ be a limit point of the sequence $\left\{x^k\right\} $, obtained by minimizing the function $f(x)$ from (\ref{f}) using the GSA-CP algorithm. Then:
\begin{enumerate}
	\item 
	if at the same time with $\overline{x}$ there exists another limit point $x'$ of the sequence $\left\{x^k\right\}$, then $f(\overline{x}) = f(x')$;
	\item
	$z^k=f(x^k) \geq f(\overline{x})$ for any  $k \geq 0$;
	\item
	If at some search step the value $\mu$ from (\ref{mu}) satisfies the condition
\begin{equation}\label{217}
  r \mu > 2^{3-1/N} L \sqrt{N+3},
\end{equation}
then $\overline{x}$ is a global minimizer of the function $f(x), x\in[0,S]$ and any global minimizer $x^\ast$ is also a limit point of the sequence $\left\{x^k\right\}$.
\end{enumerate}
\end{proposition}
\begin{proof}
This statement is proved in a similar way to the theorem on the convergence of the index method of global optimisation from \citet{Strongin2000}, taking into account the relationship (\ref{f}) of the functions $f(x)$ and $\varphi(y,u_s)$, as well as the rules (\ref{index}), specifying the indices of the trial points.
\end{proof}

@@

\subsection{Results on live hyperparameters tuning}

@@

An additional series of experiments to compare iOpt, Optuna, and HyperOpt were conducted using the JAHS-Bench-201 framework described in \citet{jahsbench}. JAHS-Bench provides convenient access to a set of tasks to determine the optimal combination of hyperparameters and neural network architecture. This tool for comparing HPO algorithms contains an extensive dataset on the quality and performance of neural networks that comprises approximately 161 million data points, each consisting of 20 performance metrics.

@@

A particular feature of JAHS-Bench is the use of surrogate models fit on performance data, covering different optimization problems (single/multiple objectives), which makes it possible to modify the neural network architecture (choosing one of the available options) as well as its hyperparameters. At the same time, by using surrogate models one can evaluate the performance of a trained classifier for a particular set of hyperparameters in a fraction of a second instead of several hours that would be required to train a real neural network. JAHS-Bench uses three popular datasets applied for classification tasks (CIFAR-10, Colorectal-Histology and Fashion MNIST) as the basis for surrogate models.

In our experiments, hyperparameter tuning using iOpt, Optuna and HyperOpt frameworks was performed on the fixed neural network architecture proposed in JAHS-Bench as the default architecture. The most relevant (from the researcher's point of view) hyperparameters were chosen for tuning: the categorical parameter of activation function selection (taking on the three possible values ReLU, Mish, Hardswish), as well as the continuous parameters Learning Rate and Weight Decay in the ranges $[10^{-3},1]$ and $[10^{-5},10^{-2}]$ respectively.

The validation accuracy metric calculated after 200 training epochs was used as the objective function. The search accuracy in the stopping criterion of all methods was set to 0.01 and the maximum number of trials was set to 200. For the algorithms from Optuna and HyperOpt, the obtained data were averaged over 10 runs. The comparison results are shown in Table \ref{tab:met-surr}. As before, the values corresponding to the best result on average are highlighted in bold and the values corresponding to the best result for the worst case are highlighted in red.

The results show the superiority of iOpt on the CIFAR-10 and Colorectal-Histology datasets. When using the Fashion MNIST dataset, the iOpt framework outperforms HyperOpt, but is inferior to Optuna. Note that when using surrogate models, the computation time of the objective function takes only fractions of a second, so the results related to the comparison of algorithm running times are not reported for this experiment.
